---
title: "Control-Flow Constructs"
author: "36-350 -- Statistical Computing"
date: "Week 2 -- Fall 2020"
output: 
  slidy_presentation: 
    font_adjustment: -1
---

Control-Flow Constructs
===

The phrase "control-flow constructs" that `R`'s creators use to denote `if` and `while` and such.

- You can find a full list of control-flow constructs by typing `?Control` in the `RStudio` console.

- The constructs are among the reserved words in `R`'s command-line parser (meaning you cannot use them as variables). You can find a list of reserved words by typing `?Reserved` in the console.

```{r error=TRUE}
xx = 2
if = 2
```

(Important note: if you want to knit an Rmd file which has a buggy code chunk, add `error=TRUE` in the first line of the chunk, e.g., <tt>```\{r error=TRUE\}</tt>).

if-else
===

Here are basic examples of the use of the `if-else` construct:
```{r}
x = 2

if ( x == 2 ) print(x)   # one line, no curly brackets

if ( x == 2 ) {          # use curly brackets to enclose multiple commands over several lines
  print(x)
}

if ( x == 2 ) {          # same idea, but now with an "else" thrown in
  print(-x)
} else {
  print(x)
}
```
(Note the `==`. `x == 2` is a relational operator, and returns `TRUE`. Use of a single `=` will lead to an exception being thrown.)

if-else: Short-Circuiting
===

An if-else construct will evaluate until the first time the quantity in parentheses evaluates to `TRUE`. Below,  $x = 2$ <i>and</i> $x > 0$, but `foo` is not printed.
```{r}
if ( x == 2 ) { 
  print(-x)
} else if ( x > 0 ) {
  print("foo")
} else {
  print("bar")
}
```

if: Logical And and Or
===

You combine conditions within an `if` statement using `&&` (for "and") or `||` (for "or").

- <b>IMPORTANT</b>: the difference between, e.g., `&&` and `&` is that the latter returns a logical vector that can have a length greater than 1 and the former returns a logical vector of length 1.

```{r}
u = c(TRUE,FALSE,TRUE)
v = c(TRUE,TRUE,FALSE)
u & v
u && v
```
`x && y` only does a comparison between the first element of `u` and `v` and ignores the subsequent elements.

Long story short: if you use `&&` in an `if` statement, make sure what is on either side of the `&&` are vectors of length one. Otherwise you may not get the behavior you expect when you run your code.

ifelse
===

A compressed version of an `if-else` construct is provided by the `ifelse()` function:
```
ifelse(test,yes,no)
```
which means "if the object test returns TRUE, do whatever is specified by yes, otherwise do whatever is specified by no." For example:
```{r}
x = 2
ifelse(x==2,TRUE,FALSE)
```
