---
title: "dplyr"
author: "36-350 -- Statistical Computing"
date: "Week 4 -- Fall 2020"
output: 
  slidy_presentation: 
    font_adjustment: -1
---

Introduction 
===

To make sense of this set of slides, think of data frames as nouns and `dplyr` functions as verbs, actions that you apply to manipulate the data frames.

Here, we will learn the following `dplyr` verbs:

- `slice()`: subset rows based on integer indexing 
- `filter()`: subset rows based on logical criteria
- `select()`: select certain columns
- `arrange()`: order rows by value of a column
- `rename()`: rename columns
- `mutate()`: create new columns
- `mutate_at()`: apply a function to given columns

Calling `dplyr` verbs always outputs a new data frame; it *does not alter* the existing data frame.

So to keep the changes, we have to reassign the data frame to be the output of the pipe, as we will see below.

Load the tidyverse
===

```{r}
# install the package if it doesn't exist...
# require() will load it if it is there as a side-effect
if ( require(tidyverse) == FALSE ) {
  install.packages("tidyverse",repos="https://cloud.r-project.org")
  suppressWarnings(library(tidyverse))
}
```

slice()
===

Use `slice()` when you want to indicate certain row numbers need to be kept:

```{r}
mtcars %>% slice(c(7,8,14:15))

# Compare with base R: mtcars[c(7,8,14:15),]
```

Slicing can also be done "negatively":

```{r}
mtcars %>% slice(-c(1:2,19:23)) %>% nrow()
```

filter()
===

Use `filter()` when you want to subset rows based on logical conditions:

```{r}
mtcars %>% filter(., (mpg >= 14 & disp >= 200) | (drat <= 3)) %>% head(., 2)

# Base R:
# head(subset(mtcars, (mpg >= 14 & disp >= 200) | (drat <= 3)), 2)
# head(mtcars[(mtcars$mpg >= 14 & mtcars$disp >= 200) | (mtcars$drat <= 3),], 2)
```

select()
===

Use `select()` when you want to pick out certain columns:

```{r}
mtcars %>% select(., cyl, disp, hp) %>% head(., 2)

# Base R:
# head(mtcars[, c("cyl", "disp", "hp")], 2)
```

Handy Arguments for select()
===

Very handy selections using `dplyr` helper functions:
```{r}
mtcars %>% select(., starts_with("d")) %>% head(., 2)

# Base R (yikes!):
# d_colnames = grep(x = colnames(mtcars), pattern = "^d")
# head(mtcars[, d_colnames], 2)
```

We can do many other things as well:
```{r}
mtcars %>% select(., ends_with('t')) %>% head(., 2)
mtcars %>% select(., ends_with('yl')) %>% head(., 2)
mtcars %>% select(., contains('ar')) %>% head(., 2)
```

arrange()
===

Use `arrange()` to order rows by values of a column:
```{r}
mtcars %>% arrange(., desc(drat)) %>% select(., mpg, disp, drat) %>% head(., 2)

# Base R:
# drat_inds = order(mtcars$drat, decreasing=TRUE)
# head(mtcars[drat_inds, c("mpg", "disp", "drat")], 2)
```

We can order by multiple columns too:
```{r}
mtcars %>% arrange(., desc(gear), desc(hp)) %>% select(., gear, hp, everything()) %>% head(., 8)
```
Here, we arrange first in order of `gear` (decreasing); then, for each unique value of `gear`, we arrange in order of `hp` (decreasing).

mutate()
===

Use `mutate()` when you want to create one or several columns:
```{r}
mtcars = mtcars %>% mutate(., hp_wt = hp/wt, mpg_wt = mpg/wt) 

# Base R:
# mtcars$hp_wt = mtcars$hp/mtcars$wt
# mtcars$mpg_wt = mtcars$mpg/mtcars$wt
```

mutate_at()
===

Use `mutate_at()` when you want to apply a function to one or several columns:

```{r}
mtcars = mtcars %>% mutate_at(., c("hp_wt", "mpg_wt"), log) 

# Base R:
# mtcars$hp_wt = log(mtcars$hp_wt)
# mtcars$mpg_wt = log(mtcars$mpg_wt)
```

**Note**: again, calling `mutate_at()` outputs a new data frame, it does not alter the given data frame, so to keep the column transformations, we have to reassign `mtcars` to be the output of the pipe!

rename()
===

Use `rename()` to easily rename columns:

```{r}
mtcars %>% rename(., hp_wt_log=hp_wt, mpg_wt_log=mpg_wt) %>% head(., 2)

# Base R:
# colnames(mtcars)[colnames(mtcars)=="hp_wt"] = "hp_wt_log"
# colnames(mtcars)[colnames(mtcars)=="mpg_wt"] = "mpg_wt_log"
# head(mtcars, 2)
```
