---
title: "Pipes"
author: "36-350 -- Statistical Computing"
date: "Week 4 -- Fall 2020"
output: 
  slidy_presentation: 
    font_adjustment: -1
---

The Pipe Operator
===

The `tidyverse`, which contain variants of split-apply-combine functions existing outside of base `R`, relies on so-called "pipes." What do we need to know about pipes?

- `tidyverse` functions perform at their best when strung together using the pipe operator.
- The pipe operator is `%>%`. **NOTE**: use `ctrl + shift + m` as RStudio shortcut.
- The pipe operator actually comes from the `magrittr` package (automatically loaded with the `tidyverse`), so it can be used on its own, completely independently.

But what does a pipe *do*? 

It takes the return value from one function call and automatically feeds it in as an input to another function.

Note that in Unix, for instance, we also have pipes, as in:
```
ls -l | grep tidy | wc -l
```

magrittr
===

Normally you would just load the entire `tidyverse`, but just to make the point that the `magrittr` package is the one with the pipe, and that the pipe can be used with non-`tidyverse` functions, we'll load `magrittr` alone.
```{r}
# install the package if it doesn't exist...
# require() will load it if it is there as a side-effect
if ( require(magrittr) == FALSE ) {
  install.packages("magrittr",repos="https://cloud.r-project.org")
  library(magrittr)
}
```

Example
===

We can write `exp(1)` with pipes as `1 %>% exp()`, and `log(exp(1))` as `1 %>% exp() %>% log()`
```{r}
exp(1)
1 %>% exp() # actually, you don't even need the parantheses if there was to be only one argument
1 %>% exp %>% log
```

How to Read a Piped Statement
===

Passing a single argument through pipes, we interpret something like:
```{r, eval=FALSE}
x %>% f %>% g %>% h
```
as `h(g(f(x)))`. So, in your mind, when you see `%>%`, you should say "and then."

To pass a single value `x` to a function that has, e.g., two arguments (`x` and `y`), you would write
```{r, eval=FALSE}
x %>% f(y) 
```
Here, `x` is assumed to be the first argument of `f()`, and `y` the second. Later we will see how to generalize this so that `x` need not be the first argument of `f()`.

Example
===

See if you can interpret the command below without running it, then run it in your `R` console to check your understanding:

```{r, eval=FALSE}
mtcars %>% head(4) %>% View
```

(Note: `mtcars` is one of the many datasets that are included in the base `R` environment. To see a list of all of the datasets, use `library(help="datasets")`.)

---

The answer: it takes the data frame `mtcars` (32 rows, 11 columns), extracts the first four rows, then passes those four rows to a viewer.

What's the "old school" (i.e., base R) way of doing this?

```{r, eval=FALSE}
View(head(mtcars,4))
```

Notice that, with pipes:

- Your code is more easily read and interpreted (arguably)
- You can run partial commands more easily

The Dot
===

The command `x %>% f(y)` can be equivalently written in *dot notation* as:
```{r, eval=FALSE}
x %>% f(.,y)
```
What's the advantage of using dots? Sometimes you want to pass in a variable as the *second* or *third* (say, not first) argument to a function, with a pipe. As in:
```{r, eval=FALSE}
x %>% f(y,.)
```
which is equivalent to evaluating `f(y,x)`.

Example
===

Again, see if you can interpret the code below without running it, then run it in your `R` console as a way to check your understanding:
```{r, eval=FALSE}
x = seq(-2*pi,2*pi,len=1000)
x %>% sin %>% abs %>% plot(x,.,typ="l")
```
