---
title: "String Basics"
author: "36-350 -- Statistical Computing"
date: "Week 3 -- Fall 2020"
output: 
  slidy_presentation: 
    font_adjustment: -1
---

What is a String?
===

A string is a sequence of characters that are bound
together, where a character is a symbol is a written language.

In `R`, a string is of mode (or class) `character` and is bounded by quotes (either single or double):
```{r}
mode("r")
class('r')
```
Two notes:

- `R` helps keep your life complicated via two similar but not identical concepts: `mode` and `class`. Feel free to lose time trying to fully internalize how they differ.

- Double quotes are preferable, because then one can use single quotes as apostrophes in strings.

Formatting Strings
===

There are special symbols that one uses for tabbing ("\\t") and for 
forcing line breaks ("\\n"):
```{r}
message = "To-do list for 350 students:\n\tHomework\n\tLabs\n\tFinal Project"
cat(message)
```

Counting Characters
===

To count the number of characters in a string, use the function `nchar()`
(as opposed to `length()`, which counts the number of elements in a vector).

To illustrate the difference, let's define a vector of strings: 
```{r}
str.vec = c("I","will","master","R")
nchar(str.vec)
length(str.vec)
```

Useful Functions for Vectors (of Strings)
===

Potentially useful functions to use with vectors that we have not seen previously include:

```{r}
head(str.vec,2)       # display/extract the first two elements
tail(str.vec,2)       #                 the last two elements
toupper(str.vec)      # change all characters to upper case
tolower(str.vec)      # change all characters to lower case
```

Substring Extraction and Replacement
===

Let's define another string vector:
```{r}
candidates = c("Trump","Clinton","Johnson","Stein") 
```
If you, for instance, want to extract the first two letters of each name,
you can use `substr()`:
```{r}
substr(candidates,1,2)
```
If you want to extract the last two letters, utilize `nchar()`, since the total number of letters in each name is different:
```{r}
substr(candidates,nchar(candidates)-1,nchar(candidates))
```

To replace the last two letters with something else,
```{r}
substr(candidates,nchar(candidates)-1,nchar(candidates)) = ":)"
candidates
```
If there is a mismatch between the number of characters to replace (here: 2) and the number of characters in the replacement string (here: 6), `R` will simply truncate. Here, that means only the first two characters in the replacement string are used.
```{r}
substr(candidates,nchar(candidates)-1,nchar(candidates)) = ":(:|:)"
candidates
```

