---
title: "Extraction/Replacement"
author: "36-350 -- Statistical Computing"
date: "Week 3 -- Fall 2020"
output: 
  slidy_presentation:
    font_adjustment: -1
---

Extraction
===

Let's say that instead of receiving an entire string that contains a matching substring, which is what you would receive if you use `grep()` with `value=TRUE`, you just want that substring. To get just the substring, use a combination of `regexpr()` and `regmatches()`:

- `regexpr()` returns the location of the *first* match in the target string (or -1 if no match is found); and

- `regmatches()` takes the output of `regexpr()` and returns the matching substring.

Extraction Example
===

```{r}
str.vec = c("abccc","ababc","cabcc","ccabb","ccccc")
reg.exp = regexpr("ab+",str.vec,useBytes=TRUE)
reg.exp
```

The first line of output tells you *where* in the string the match occurs, and that the `match.length` attribute tells you the length of the matching substring in characters.

```{r}
regmatches(str.vec,reg.exp)
```

Replacement
===

In a previous example, I showed how one outputs the matching substrings:
```{r}
str.vec = c("abccc","ababc","cabcc","ccabb","ccccc")
reg.exp = regexpr("ab+",str.vec,useBytes=TRUE)
regmatches(str.vec,reg.exp)
```

But you can also use `regmatches()` to specify a *replacement* for these substrings:
```{r}
regmatches(str.vec,reg.exp) = "xy"
str.vec
```

Notice that in the second string, only the first instance of "ab" is replaced. Again, as we have been using it up to now, `regexpr()` will only return the *first* matching substring. To get all the matching substrings in a given string, use `gregexpr()`.

gregexpr()
===

```{r}
str.vec  = c("abab","abAB","ABAB","ccAbCC")
greg.exp = gregexpr("ab|AB",str.vec)
regmatches(str.vec,greg.exp)                # returns a list, one element per input string
```

```{r}
regmatches(str.vec,greg.exp) = "xy"
str.vec
```

Alternative Replacement: (g)sub()
===

```{r}
str.vec = c("abab","abAB","ABAB","ccAbCC")
sub("ab|AB","xy",str.vec)                   # sub(): replace first occurrence of matching substring
str.vec
```
Note that unlike before, the original string vector is not itself changed! If you need, e.g., to compare how your initial and final strings appear, `sub()` or `gsub()` is the way to go. The difference: `sub()` replaces the first occurrence of the matching substring, while `gsub()` replaces all occurrences of the matching substring.
```{r}
str.vec = c("abab","abAB","ABAB","ccAbCC")
gsub.out = gsub("ab|AB","xy",str.vec)
gsub.out
```

